How functional traits influence plant growth and shade tolerance across the life-cycle

### 2018-05-17 ms/ms.tex
Change labelling of equations in Figures and Tables
journal requested using numbers 1-n for equations in Fig 1 and Table 4

In the main text I have hard-coded the numbers

% Enable equation numbers in Table
% Following http://tex.stackexchange.com/questions/62974/numbering-equations-in-tabular-environment
\newcommand{\eqncounter}[1]{%
\mbox{}\refstepcounter{equation}%
$[\theequation]$%
\label{#1}
}

@@

Earlier studies on seedlings related relative growth rate in mass to {\lma} \citep{Blackman-1919, Lambers-1992, Cornelissen-1996, Wright-2000}, using a model that can be derived from eqn \ref{eq:dbdt} as a special case of the more-extended model described below. Ignoring all turnover terms, as well as the respiration terms for non-leaf tissues in eqn \ref{eq:dbdt} and reproduction, biomass growth becomes a linear function of leaf area, making relative growth rate in mass:
\begin{equation}\label{eq:RGR}
\frac{{\rm d}B}{{\rm d}t}\frac{1}{M_{\rm a}}  \approx P_{\rm net} \times \phi^{-1} \times \frac{M_{\rm l}}{M_{\rm a}}, \end{equation}
where $P_{\rm net} = \alpha_{\rm bio} \, \alpha_{\rm y}\,(\bar{p}(E) - r_{\rm l})$, and $\phi$ is {\lma}. Although eqn \ref{eq:RGR} captures patterns of growth in seedlings in relation to {\lma} \citep{Wright-2000}, this approximation does not easily extend to the variables that are routinely collected for large trees such as height or stem diameter. The derivations below clarify these links.

@@

\refstepcounter{equation}\label{eq:dbdt_2}
\refstepcounter{equation}\label{eq:dMtdt}
\refstepcounter{equation}\label{eq:daldmt}
\refstepcounter{equation}\label{eq:daldt}
\refstepcounter{equation}\label{eq:dhdt}
\refstepcounter{equation}\label{eq:dast}
\refstepcounter{equation}\label{eq:dDdt}

% Update equation counter, following inclusion of figure 1
% \setcounter{equation}{9}

To model growth in plant height ($H$), leaf area ($A_{\rm l}$), basal stem area ($ A_{\rm st}$), standing mass ($M_{\rm t}$) or stem diameter ($D$) requires that we account not just for biomass growth, but also for the costs of building new tissues and allocation to reproduction. Mathematically, these growth rates can be decomposed into a product of physiologically relevant terms \citep{Falster-2011, Gibert-2016} (Fig. \ref{fig:conceptual}b, eqns \ref{eq:dMtdt}-\ref{eq:dDdt}). While other decompositions are possible, the approach used here arises from the idea that multiple aspects of function and allocation decisions are made in relation to the amount of leaf area on a plant  \citep{Falster-2011}. Growth rates in plant weight ($\frac{\ud M_{\rm t} }{\ud t}$; eqn \ref{eq:dMtdt}), total leaf area ($\frac{\ud A_{\rm l}}{\ud t}$; eqn \ref{eq:daldt}), height ($\frac{\ud H}{\ud t}$; eqn \ref{eq:dhdt}), stem basal area ($\frac{\ud  A_{\rm st}}{\ud t}$; eqn \ref{eq:dast}), and stem diameter ($\frac{\ud D}{\ud t}$; eqn \ref{eq:dDdt}) therefore share several terms. Many of these terms also vary intrinsically with size, as shown in the insets in Fig. \ref{fig:conceptual}b.

The growth rate of all size metrics in eqns \ref{eq:dMtdt}-\ref{eq:dDdt} depends on the product of biomass growth $\frac{{\rm d}B}{{\rm d}t}$ (from eqn \ref{eq:dbdt}) and the fraction of biomass allocated to growth of the plant, $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$, which varies from 0-1. The remaining $1- \frac{{\rm d}M_{\rm a}}{{\rm d}B}$ fraction of mass produced is allocated to reproduction. In plants, $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ starts high, 1.0 for seedlings, and then decreases through ontogeny, potentially to zero in fully mature plants \citep{Wenk-2015}. Note also that $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ is the allocation of biomass \emph{after} replacing parts lost due to turnover. So a plant with $\frac{{\rm d}M_{\rm a}}{{\rm d}B}=0$ will continue to produce some new leaves and increase in stem diameter, even if the net amount of live mass $M_{\rm a}$ is not increasing.

@@

\caption{Conceptual framework showing how shade-tolerance (as measured by \textsc{wplcp}) is affected by plant size and traits, adapted from \citep{Givnish-1988}. From eqn \ref{eq:wplcp}, \textsc{wplcp} is the canopy openness $E$ where the photosynthetic income per unit leaf area (dashed line) intersects the total respiration and turnover costs per unit leaf area (solid black lines). Colors indicate the contributions from different tissues to total costs. Because the costs of stem tissues increase with height, \textsc{wplcp} also increases. Traits {\wood}, {\nitrogen}, and {\lma} can alter \textsc{wplcp} by shifting either income or cost of different tissues (Table \ref{tab:responses}).
\label{fig:wplcp_idea}}
\end{figure}

The growth rate in the total standing mass of the plant (eqn \ref{eq:dMtdt}) is then the sum of heartwood formation (=sapwood turnover) and any increment in live mass.

The remaining growth rates (eqns \ref{eq:daldt}-\ref{eq:dDdt}) all depend on another variable, $\frac{{\rm d}A_{\rm l}} {{\rm d}M_{\rm a}}$, that accounts for the marginal cost of deploying an additional unit of leaf area, including construction of the leaf itself and supporting  bark, sapwood, and roots (eqn \ref{eq:daldmt}). The inverse of this term, $\frac{{\rm d}M_{\rm a} }{{\rm d}A_{\rm l}}$, is the whole plant construction cost per unit leaf area, which can be further decomposed as a sum of construction cost per unit leaf area for different tissues, with one of these being the trait $\phi = \frac{{\rm d}M_{\rm l} }{{\rm d}A_{\rm l}}$ (eqn \ref{eq:daldmt}).

The rate of height growth (eqn \ref{eq:dhdt}) depends on an additional term, $\frac{{\rm d}H }{{\rm d}A_{\rm l}}$: the growth in plant height per unit growth in leaf area. This variable accounts for the architectural strategy of the plant (\citep{Falster-2011}). Some species tend to leaf out more than grow tall, while other species emphasise vertical extension \citep{Poorter-2006}.

The rate of stem-basal-area growth (eqn \ref{eq:dast}) can be expressed as the sum of increments in sapwood, bark, and heartwood areas ($A_{\rm s}, A_{\rm b}, A_{\rm h}$ respectively): $\frac{{\rm d}A_{\rm st}}{{\rm d}t}= \frac{{\rm d}A_{\rm b}}{{\rm d}t} + \frac{{\rm d}A_{\rm s}}{{\rm d}t} + \frac{{\rm d}A_{\rm h}}{{\rm d}t}$. These in turn are related to ratios of sapwood and bark area per leaf area, and sapwood turnover (see eqn \ref{eq:dast}).

Finally, the rate of stem-diameter growth (eqn \ref{eq:dDdt}) is given by a geometric relationship between stem diameter ($D$) and stem area ($A_{\rm st}$). We make no assumptions about the relationship of stem diameter to height or leaf area: these arise as emergent properties, via integration of stem turnover (eqns \ref{eq:ha}-\ref{eq:htm}).

@@

Eqn \ref{eq:dbdt} can also be rearranged to obtain a measure of shade tolerance: the ``whole-plant-light-compensation point'' \citep[{\wplcp}][]{Givnish-1988, Baltzer-2007, Lusk-2013} . In general, photosynthesis rate per leaf area $\bar{p}$ increases with canopy openness, $E$. The {\wplcp} is then the value $E=E^*$ where photosynthetic gains balance the costs of tissue turnover and respiration, i.e. $\frac{{\rm d}B}{{\rm d}t}(E^*)=0$. From eqn \ref{eq:dbdt}, this occurs when
\begin{equation}\label{eq:wplcp}
\bar{p}(E^*) =\frac{\sum_{i={\rm l,b,s,r}} \, M_{\rm i} \left(\frac{k_{\rm i}}{y} + r_{\rm i}\right)}{A_{\rm l}}.
\end{equation}
Graphically, the {\wplcp} occurs at the point where the photosynthetic production (per unit leaf area) line intersects with the sum of maintenance and respiration costs (per unit leaf area) for each tissue (Fig. \ref{fig:wplcp_idea}). Traits influence {\wplcp} when they affect either photosynthesis, respiration, or turnover. Also, as the amount of stem support increases with plant height, {\wplcp} also increases with height \citep{Givnish-1988} (Fig. \ref{fig:wplcp_idea}).

@@

Since eqns \ref{eq:dMtdt}-\ref{eq:wplcp} are derived using standard rules of addition, multiplication, and differentiation, they may apply to a variety of growth model where biomass allocation is important. However, to make explicit predictions requires that we quantify all the terms in Fig. \ref{fig:conceptual}b via an explicit model of plant construction and function.

The {\plant} package adopts a model of plant construction and function that can be considered a first-order functional-balance or function-equilibrium model, similar to those implemented by \citep{Makela-1997} and \citep{Moorcroft-2001}. We could also call it ``isometric'', because the assumptions see area-based metrics scaling to the first-power of other area-based metrics, and to the square-power of length-based metrics, such as height \citep{Huxley-1932}. Table \ref{tab:allometry} provides key equations (see \citep{Falster-2016} for full derivation). In particular, we assume that as a plant grows:
1) Its height scales to the 0.5 power of its leaf area (eqn \ref{eq:ha});
2) The cross-sectional area of sapwood in the stem is proportional to its leaf area (eqn \ref{eq:as});
3) The cross-sectional area of bark and phloem in the stem  is proportional to its leaf area (eqn \ref{eq:ab});
4) The cross-sectional area of root surface area and therefore mass is proportional to its leaf area (eqn \ref{eq:mr}); and
5) The vertical distribution of leaf within the plant's canopy, relative to the plant's height, remains constant.
Assumption 1 accounts for the architectural layout of the plant. Assumptions 2-4 are realisations of the pipe model \citep{Shinozaki-1964}, whereby the cross-sectional area of conducting tissues are proportional to leaf area. To describe the vertical distribution of leaf area within the canopy of an individual plant (assumption 5), we use the model of \citep{Yokozawa-1995}, which can account for a variety of canopy profiles through a single parameter $\eta_c$, varying from 0-1 (for details, see SI).

Combined, the functional-balance assumptions from Table \ref{tab:allometry}a lead directly to equations describing the mass of sapwood and bark in relation to leaf, and the amount of leaf in relation to height (Table \ref{tab:allometry}b). These eqns aso predict that the amount of live stem tissue supporting each unit of leaf area should increase linearly with height,
\begin{equation}\label{eq:msml}
\frac{M_{\rm b} + M_{\rm s}}{A_{\rm l}} = (1+ \alpha_{\rm b1}) \,\theta \, \rho \, \eta_c \, H.
\end{equation}
Substituting from Table \ref{tab:allometry} into eqns \ref{eq:dhdt}, \ref{eq:dast}, \ref{eq:dDdt} then gives all the necessary terms needed to implement the growth model described in Fig. \ref{fig:conceptual}.

\begin{table*}
\caption{Equations for a ``functional-balance'' model of plant construction. See Table \ref{tab:definitions} for a list of variable names and definitions.}
\centering
  \begin{tabular}{p{0.1cm}p{2.5cm}p{3cm}p{4cm}p{2.5cm}p{0.75cm}}
  \toprule
  & Variable & Function & Marginal cost & Growth rate & Eqn\\
  \midrule
  \multicolumn{6}{l}{\textbf{a) Functional-balance assumptions}} \\
  & Height &
    $H = \alpha_{\rm l1} \, A_{\rm l}^{0.5}$ &
    $\frac{{\rm d}H}{{\rm d}A_{\rm l}} = 0.5 \alpha_{\rm l1} A_{\rm l}^{-0.5}$ &
    $\frac{{\rm d}H}{{\rm d}t}  = \frac{{\rm d}H}{{\rm d}A_{\rm l}} \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ &  \eqncounter{eq:ha}\\
  & Sapwood area &
    $A_{\rm s} = \theta \, A_{\rm l}$ &
    $\frac{{\rm d}A_{\rm s}}{{\rm d} A_{\rm l}} = \theta$ &
    $\frac{{\rm d}A_{\rm s}}{{\rm d}t}  =\frac{{\rm d}A_{\rm s}}{{\rm d} A_{\rm l}} \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & \eqncounter{eq:as}\\
  & Bark area &
    $A_{\rm b} = b \, \theta \, A_{\rm l}$ &
    $\frac{{\rm d}A_{\rm b}}{{\rm d} A_{\rm l}} = b \, \theta$ &
    $\frac{{\rm d}A_{\rm b}}{{\rm d}t} = \frac{{\rm d}A_{\rm b}}{{\rm d} A_{\rm l}} \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & \eqncounter{eq:ab}\\
  & Root mass &
    $M_{\rm r} = \alpha_{\rm r1} \, A_{\rm l}$ &
    $\frac{{\rm d}M_{\rm r}}{{\rm d}A_{\rm l}} = \alpha_{\rm r1}$  &
    $\frac{{\rm d}M_{\rm r}}{{\rm d}t}  = \frac{{\rm d}M_{\rm r}}{{\rm d}A_{\rm l}}  \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & \eqncounter{eq:mr}\\
  & Heartwood area & & &
    $\frac{{\rm d}A_{\rm h}}{{\rm d} t}  = k_{s} A_{\rm h} $ & \eqncounter{eq:dah}\\
  \multicolumn{6}{l}{\textbf{b) Derived quantities}} \\
  & Leaf mass &
    $M_{\rm l} = \phi \, A_{\rm l} $ &
    $\frac{{\rm d}M_{\rm l}}{{\rm d}A_{\rm l}} = \phi$ &
    $\frac{{\rm d}M_{\rm l}}{{\rm d}t}  = \frac{{\rm d}M_{\rm l}}{{\rm d}A_{\rm l}}  \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & \eqncounter{eq:ml}\\
  & Sapwood mass &
    $M_{\rm s} = \theta \, \rho \, \eta_c \, A_{\rm l} \, H $ &
    $\frac{{\rm d}M_{\rm s}}{{\rm d}A_{\rm l}} = \theta\, \rho\, \eta_c\, \big( H + A_{\rm l}\, \frac{{\rm d}H}{{\rm d}A_{\rm l}} \big)$ &
    $\frac{{\rm d}M_{\rm s}}{{\rm d}t}  = \frac{{\rm d}M_{\rm s}}{{\rm d}A_{\rm l}} \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & \eqncounter{eq:ms2}\\
  & Bark mass &
    $M_{\rm b} = b\, \theta \, \rho \, \eta_c \, A_{\rm l} \, H $ &
    $\frac{{\rm d}M_{\rm b}}{{\rm d}A_{\rm l}} = b \, \theta \, \rho \, \eta_c\big( H + A_{\rm l} \, \frac{{\rm d}H}{{\rm d}A_{\rm l}} \big)$ &
    $\frac{{\rm d}M_{\rm b}}{{\rm d}t}  = \frac{{\rm d}M_{\rm b}}{{\rm d}A_{\rm l}} \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & \eqncounter{eq:mb}\\
  & Heartwood area &
    $A_{\rm h} = \int_0^t \frac{{\rm d}A_{\rm h}}{{\rm d}t}(t^\prime) \, dt^\prime$ &
     &
    $\frac{{\rm d}A_{\rm h}}{{\rm d}t} = k_{\rm s} \, A_{\rm s}$ & \eqncounter{eq:hta} \\
  & Heartwood mass &
    $M_{\rm h} = \int_0^t \frac{{\rm d}M_{\rm h}}{{\rm d}t}(t^\prime) \, dt^\prime$ &
     &
    $\frac{{\rm d}M_{\rm h}}{{\rm d}t} = k_{\rm s} \, M_{\rm s}$ & \eqncounter{eq:htm} \\
  \bottomrule
\end{tabular}
\addtabletext{\\The first column of \textbf{a)} provides cores assumptions between various size metrics and leaf area. Eqns in the middle and right columns of \textbf{a)} and in \textbf{b)} can be derived from these assumptions. The column ``Eqn'' indicates equation numbers referred to in the main text.

}
\label{tab:allometry}
\end{table*}

@@

Height at maturation: {\hmat} moderates an inevitable trade-off between mass invested reproduction and growing the plant, i.e.~ $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ in eqns \ref{eq:dhdt} and \ref{eq:dast}. We assume the amount allocated to vegetative growth starts high and then decreases once $H > H_{{\rm mat}}$  (Fig. \ref{fig:conceptual}b). The exact nature of the function is non-critical, what is important is that plants shift from a period of investing mainly in vegetative growth to investing mainly in reproduction. Here we let the fraction of mass allocated to reproduction vary with $H$ via the function $r_{\rm r1}  \left(1.0 + \exp\left(r_{\rm r2} \left(1 - H / H_{{\rm mat}}\right)\right)\right)^{-1}$, where $r_{\rm r1}$ is the maximum possible allocation (0-1) and $r_{\rm r2}$ determines the sharpness of the transition.

>>>>

@@

Earlier studies on seedlings related relative growth rate in mass to {\lma} \citep{Blackman-1919, Lambers-1992, Cornelissen-1996, Wright-2000}, using a model that can be derived from Eq. \ref{eq:dbdt} as a special case of the more-extended model described below. Ignoring all turnover terms, as well as the respiration terms for non-leaf tissues in Eq. \ref{eq:dbdt} and reproduction, biomass growth becomes a linear function of leaf area, making relative growth rate in mass:
\begin{equation}\label{eq:RGR}
\frac{{\rm d}B}{{\rm d}t}\frac{1}{M_{\rm a}}  \approx P_{\rm net} \times \phi^{-1} \times \frac{M_{\rm l}}{M_{\rm a}}, \end{equation}
where $P_{\rm net} = \alpha_{\rm bio} \, \alpha_{\rm y}\,(\bar{p}(E) - r_{\rm l})$, and $\phi$ is {\lma}. Although Eq. \ref{eq:RGR} captures patterns of growth in seedlings in relation to {\lma} \citep{Wright-2000}, this approximation does not easily extend to the variables that are routinely collected for large trees such as height or stem diameter. The derivations below clarify these links.

@@

To model growth in plant height ($H$), leaf area ($A_{\rm l}$), basal stem area ($ A_{\rm st}$), standing mass ($M_{\rm t}$) or stem diameter ($D$) requires that we account not just for biomass growth, but also for the costs of building new tissues and allocation to reproduction. Mathematically, these growth rates can be decomposed into a product of physiologically relevant terms \citep{Falster-2011, Gibert-2016} (Fig. \ref{fig:conceptual}b, Eqs. {\color{pnasbluetext}2}-{\color{pnasbluetext}7}). While other decompositions are possible, the approach used here arises from the idea that multiple aspects of function and allocation decisions are made in relation to the amount of leaf area on a plant  \citep{Falster-2011}. Growth rates in plant weight ($\frac{\ud M_{\rm t} }{\ud t}$; Fig. \ref{fig:conceptual}b, Eqs. {\color{pnasbluetext}2}), total leaf area ($\frac{\ud A_{\rm l}}{\ud t}$; Fig. \ref{fig:conceptual}b, Eq. {\color{pnasbluetext}4}), height ($\frac{\ud H}{\ud t}$; Fig. \ref{fig:conceptual}b, Eq. {\color{pnasbluetext}5}), stem basal area ($\frac{\ud  A_{\rm st}}{\ud t}$; Fig. \ref{fig:conceptual}b, Eq. {\color{pnasbluetext}6}), and stem diameter ($\frac{\ud D}{\ud t}$; Fig. \ref{fig:conceptual}b, Eq. {\color{pnasbluetext}7}) therefore share several terms. Many of these terms also vary intrinsically with size, as shown in the insets in Fig. \ref{fig:conceptual}b.

The growth rates of all size metrics (Fig. \ref{fig:conceptual}b, Eqs. {\color{pnasbluetext}2}-{\color{pnasbluetext}7}) depend on the product of biomass growth $\frac{{\rm d}B}{{\rm d}t}$ (from Eq. \ref{eq:dbdt}) and the fraction of biomass allocated to growth of the plant, $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$, which varies from 0-1. The remaining $1- \frac{{\rm d}M_{\rm a}}{{\rm d}B}$ fraction of mass produced is allocated to reproduction. In plants, $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ starts high, 1.0 for seedlings, and then decreases through ontogeny, potentially to zero in fully mature plants \citep{Wenk-2015}. Note also that $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ is the allocation of biomass \emph{after} replacing parts lost due to turnover. So a plant with $\frac{{\rm d}M_{\rm a}}{{\rm d}B}=0$ will continue to produce some new leaves and increase in stem diameter, even if the net amount of live mass $M_{\rm a}$ is not increasing.

@@

\caption{Conceptual framework showing how shade-tolerance (as measured by \textsc{wplcp}) is affected by plant size and traits, adapted from \citep{Givnish-1988}. From Eq. \ref{eq:wplcp}, \textsc{wplcp} is the canopy openness $E$ where the photosynthetic income per unit leaf area (dashed line) intersects the total respiration and turnover costs per unit leaf area (solid black lines). Colors indicate the contributions from different tissues to total costs. Because the costs of stem tissues increase with height, \textsc{wplcp} also increases. Traits {\wood}, {\nitrogen}, and {\lma} can alter \textsc{wplcp} by shifting either income or cost of different tissues (Table \ref{tab:responses}).
\label{fig:wplcp_idea}}
\end{figure}

The growth rate in the total standing mass of the plant (Fig. \ref{fig:conceptual}b, Eqs. {\color{pnasbluetext}2}) is then the sum of heartwood formation (=sapwood turnover) and any increment in live mass.

The remaining growth rates (Fig. \ref{fig:conceptual}b, Eqs. {\color{pnasbluetext}4}-{\color{pnasbluetext}7}) all depend on another variable, $\frac{{\rm d}A_{\rm l}} {{\rm d}M_{\rm a}}$, that accounts for the marginal cost of deploying an additional unit of leaf area, including construction of the leaf itself and supporting  bark, sapwood, and roots (Fig. \ref{fig:conceptual}b, Eq. {\color{pnasbluetext}3}). The inverse of this term, $\frac{{\rm d}M_{\rm a} }{{\rm d}A_{\rm l}}$, is the whole plant construction cost per unit leaf area, which can be further decomposed as a sum of construction cost per unit leaf area for different tissues, with one of these being the trait $\phi = \frac{{\rm d}M_{\rm l} }{{\rm d}A_{\rm l}}$ (Fig. \ref{fig:conceptual}b, Eq. {\color{pnasbluetext}3}).

The rate of height growth (Fig. \ref{fig:conceptual}b, Eq. {\color{pnasbluetext}5}) depends on an additional term, $\frac{{\rm d}H }{{\rm d}A_{\rm l}}$: the growth in plant height per unit growth in leaf area. This variable accounts for the architectural strategy of the plant (\citep{Falster-2011}). Some species tend to leaf out more than grow tall, while other species emphasise vertical extension \citep{Poorter-2006}.

The rate of stem-basal-area growth (Fig. \ref{fig:conceptual}b, Eq. {\color{pnasbluetext}6}) can be expressed as the sum of increments in sapwood, bark, and heartwood areas ($A_{\rm s}, A_{\rm b}, A_{\rm h}$ respectively): $\frac{{\rm d}A_{\rm st}}{{\rm d}t}= \frac{{\rm d}A_{\rm b}}{{\rm d}t} + \frac{{\rm d}A_{\rm s}}{{\rm d}t} + \frac{{\rm d}A_{\rm h}}{{\rm d}t}$. These in turn are related to ratios of sapwood and bark area per leaf area, and sapwood turnover (see Fig. \ref{fig:conceptual}b, Eq. {\color{pnasbluetext}6}).

Finally, the rate of stem-diameter growth (Fig. \ref{fig:conceptual}b, Eq. {\color{pnasbluetext}7}) is given by a geometric relationship between stem diameter ($D$) and stem area ($A_{\rm st}$). We make no assumptions about the relationship of stem diameter to height or leaf area: these arise as emergent properties, via integration of stem turnover (Table \ref{tab:allometry}, Eqs. {\color{pnasbluetext}9}-{\color{pnasbluetext}10}).

@@

Eq. \ref{eq:dbdt} can also be rearranged to obtain a measure of shade tolerance: the ``whole-plant-light-compensation point'' \citep[{\wplcp}][]{Givnish-1988, Baltzer-2007, Lusk-2013} . In general, photosynthesis rate per leaf area $\bar{p}$ increases with canopy openness, $E$. The {\wplcp} is then the value $E=E^*$ where photosynthetic gains balance the costs of tissue turnover and respiration, i.e. $\frac{{\rm d}B}{{\rm d}t}(E^*)=0$. From Eq. \ref{eq:dbdt}, this occurs when
\begin{equation}\label{eq:wplcp}
\bar{p}(E^*) =\frac{\sum_{i={\rm l,b,s,r}} \, M_{\rm i} \left(\frac{k_{\rm i}}{y} + r_{\rm i}\right)}{A_{\rm l}}.
\end{equation}
Graphically, the {\wplcp} occurs at the point where the photosynthetic production (per unit leaf area) line intersects with the sum of maintenance and respiration costs (per unit leaf area) for each tissue (Fig. \ref{fig:wplcp_idea}). Traits influence {\wplcp} when they affect either photosynthesis, respiration, or turnover. Also, as the amount of stem support increases with plant height, {\wplcp} also increases with height \citep{Givnish-1988} (Fig. \ref{fig:wplcp_idea}).

@@

Since Eqs. in Fig. \ref{fig:conceptual}b and Eq. \ref{eq:wplcp} are derived using standard rules of addition, multiplication, and differentiation, they may apply to a variety of growth model where biomass allocation is important. However, to make explicit predictions requires that we quantify all the terms in Fig. \ref{fig:conceptual}b via an explicit model of plant construction and function.

The {\plant} package adopts a model of plant construction and function that can be considered a first-order functional-balance or function-equilibrium model, similar to those implemented by \citep{Makela-1997} and \citep{Moorcroft-2001}. We could also call it ``isometric'', because the assumptions see area-based metrics scaling to the first-power of other area-based metrics, and to the square-power of length-based metrics, such as height \citep{Huxley-1932}. Table \ref{tab:allometry} provides key equations (see \citep{Falster-2016} for full derivation). In particular, we assume that as a plant grows:
1) Its height scales to the 0.5 power of its leaf area (Table \ref{tab:allometry}, Eq. {\color{pnasbluetext}1});
2) The cross-sectional area of sapwood in the stem is proportional to its leaf area (Table \ref{tab:allometry}, Eq. {\color{pnasbluetext}2});
3) The cross-sectional area of bark and phloem in the stem  is proportional to its leaf area (Table \ref{tab:allometry}, Eq. {\color{pnasbluetext}3});
4) The cross-sectional area of root surface area and therefore mass is proportional to its leaf area (Table \ref{tab:allometry}, Eq. {\color{pnasbluetext}4}); and
5) The vertical distribution of leaf within the plant's canopy, relative to the plant's height, remains constant.
Assumption 1 accounts for the architectural layout of the plant. Assumptions 2-4 are realisations of the pipe model \citep{Shinozaki-1964}, whereby the cross-sectional area of conducting tissues are proportional to leaf area. To describe the vertical distribution of leaf area within the canopy of an individual plant (assumption 5), we use the model of \citep{Yokozawa-1995}, which can account for a variety of canopy profiles through a single parameter $\eta_c$, varying from 0-1 (for details, see SI).

Combined, the functional-balance assumptions from Table \ref{tab:allometry}a lead directly to equations describing the mass of sapwood and bark in relation to leaf, and the amount of leaf in relation to height (Table \ref{tab:allometry}b). These equations also predict that the amount of live stem tissue supporting each unit of leaf area should increase linearly with height,
\begin{equation}\label{eq:msml}
\frac{M_{\rm b} + M_{\rm s}}{A_{\rm l}} = (1+ \alpha_{\rm b1}) \,\theta \, \rho \, \eta_c \, H.
\end{equation}
Substituting from Table \ref{tab:allometry} Eqs. {\color{pnasbluetext}1}-{\color{pnasbluetext}10} into Fig. \ref{fig:conceptual}b Eqs. {\color{pnasbluetext}5}-{\color{pnasbluetext}7} then gives all the necessary terms needed to implement the growth model described in Fig. \ref{fig:conceptual}b.

\begin{table*}
\caption{Equations for a ``functional-balance'' model of plant construction. See Table \ref{tab:definitions} for a list of variable names and definitions.}
\centering
  \begin{tabular}{p{0.1cm}p{2.5cm}p{3cm}p{4cm}p{2.5cm}p{0.75cm}}
  \toprule
  & Variable & Function & Marginal cost & Growth rate & Eq.\\
  \midrule
  \multicolumn{6}{l}{\textbf{a) Functional-balance assumptions}} \\
  & Height &
    $H = \alpha_{\rm l1} \, A_{\rm l}^{0.5}$ &
    $\frac{{\rm d}H}{{\rm d}A_{\rm l}} = 0.5 \alpha_{\rm l1} A_{\rm l}^{-0.5}$ &
    $\frac{{\rm d}H}{{\rm d}t}  = \frac{{\rm d}H}{{\rm d}A_{\rm l}} \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ &  [1]\\
  & Sapwood area &
    $A_{\rm s} = \theta \, A_{\rm l}$ &
    $\frac{{\rm d}A_{\rm s}}{{\rm d} A_{\rm l}} = \theta$ &
    $\frac{{\rm d}A_{\rm s}}{{\rm d}t}  =\frac{{\rm d}A_{\rm s}}{{\rm d} A_{\rm l}} \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & [2]\\
  & Bark area &
    $A_{\rm b} = b \, \theta \, A_{\rm l}$ &
    $\frac{{\rm d}A_{\rm b}}{{\rm d} A_{\rm l}} = b \, \theta$ &
    $\frac{{\rm d}A_{\rm b}}{{\rm d}t} = \frac{{\rm d}A_{\rm b}}{{\rm d} A_{\rm l}} \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & [3]\\
  & Root mass &
    $M_{\rm r} = \alpha_{\rm r1} \, A_{\rm l}$ &
    $\frac{{\rm d}M_{\rm r}}{{\rm d}A_{\rm l}} = \alpha_{\rm r1}$  &
    $\frac{{\rm d}M_{\rm r}}{{\rm d}t}  = \frac{{\rm d}M_{\rm r}}{{\rm d}A_{\rm l}}  \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & [4]\\
  & Heartwood area & & &
    $\frac{{\rm d}A_{\rm h}}{{\rm d} t}  = k_{s} A_{\rm h} $ & [5]\\
  \multicolumn{6}{l}{\textbf{b) Derived quantities}} \\
  & Leaf mass &
    $M_{\rm l} = \phi \, A_{\rm l} $ &
    $\frac{{\rm d}M_{\rm l}}{{\rm d}A_{\rm l}} = \phi$ &
    $\frac{{\rm d}M_{\rm l}}{{\rm d}t}  = \frac{{\rm d}M_{\rm l}}{{\rm d}A_{\rm l}}  \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & [6]\\
  & Sapwood mass &
    $M_{\rm s} = \theta \, \rho \, \eta_c \, A_{\rm l} \, H $ &
    $\frac{{\rm d}M_{\rm s}}{{\rm d}A_{\rm l}} = \theta\, \rho\, \eta_c\, \big( H + A_{\rm l}\, \frac{{\rm d}H}{{\rm d}A_{\rm l}} \big)$ &
    $\frac{{\rm d}M_{\rm s}}{{\rm d}t}  = \frac{{\rm d}M_{\rm s}}{{\rm d}A_{\rm l}} \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & [7]\\
  & Bark mass &
    $M_{\rm b} = b\, \theta \, \rho \, \eta_c \, A_{\rm l} \, H $ &
    $\frac{{\rm d}M_{\rm b}}{{\rm d}A_{\rm l}} = b \, \theta \, \rho \, \eta_c\big( H + A_{\rm l} \, \frac{{\rm d}H}{{\rm d}A_{\rm l}} \big)$ &
    $\frac{{\rm d}M_{\rm b}}{{\rm d}t}  = \frac{{\rm d}M_{\rm b}}{{\rm d}A_{\rm l}} \, \frac{{\rm d}A_{\rm l}}{{\rm d}t}$ & [8]\\
  & Heartwood area &
    $A_{\rm h} = \int_0^t \frac{{\rm d}A_{\rm h}}{{\rm d}t}(t^\prime) \, dt^\prime$ &
     &
    $\frac{{\rm d}A_{\rm h}}{{\rm d}t} = k_{\rm s} \, A_{\rm s}$ & [9] \\
  & Heartwood mass &
    $M_{\rm h} = \int_0^t \frac{{\rm d}M_{\rm h}}{{\rm d}t}(t^\prime) \, dt^\prime$ &
     &
    $\frac{{\rm d}M_{\rm h}}{{\rm d}t} = k_{\rm s} \, M_{\rm s}$ & [10]\\
  \bottomrule
\end{tabular}
\addtabletext{\\The first column of \textbf{a)} provides cores assumptions between various size metrics and leaf area. Eqs. in the middle and right columns of \textbf{a)} and in \textbf{b)} can be derived from these assumptions.
}
\label{tab:allometry}
\end{table*}

@@

Height at maturation: {\hmat} moderates an inevitable trade-off between mass invested in reproduction and in growing the plant, i.e.~ $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ in Fig. \ref{fig:conceptual}b Eq. {\color{pnasbluetext}2}-{\color{pnasbluetext}7}. We assume the amount allocated to vegetative growth starts high and then decreases once $H > H_{{\rm mat}}$  (Fig. \ref{fig:conceptual}b). The exact nature of the function is non-critical, what is important is that plants shift from a period of investing mainly in vegetative growth to investing mainly in reproduction. Here we let the fraction of mass allocated to reproduction vary with $H$ via the function $r_{\rm r1}  \left(1.0 + \exp\left(r_{\rm r2} \left(1 - H / H_{{\rm mat}}\right)\right)\right)^{-1}$, where $r_{\rm r1}$ is the maximum possible allocation (0-1) and $r_{\rm r2}$ determines the sharpness of the transition.
